\large End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK

### 2020-05-29 doc/2020-SEFM/paper.tex
Small changes to the Salty in SPARK paper

SPARK, which is based on the Ada programming language, is both a programming language 
with a specification language and an associated verification toolset \cite{hoang2015spark}.
% SPARK translates code to the Why language for processing by the Why3 theorem prover, 
% and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
For example, if one were to translate source code originally written in Ada to SPARK, 
one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
Formal verification of user-specified properties also requires manually writing specifications at the level of the source code in the 
form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by individual subprograms.
There are different levels to which one can use SPARK to verify a program \cite{moy2019climbing}. 
Generally, the level of verification performed during development is incremental, going from lowest to highest. 
Verification at the lowest level, colloquially referred to as ``stone level,'' is achieved when code is accepted by SPARK, 
since the SPARK language has stricter legality rules than Ada. 
Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
flow analysis checks for errors such as reads of uninitialized data or violations of user-specified data flow contracts. 
Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
e.g. division by zero or numeric overflow/underflow. 
Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
type invariants and subprogram preconditions and postconditions; 
however, at this level, the specifications only partially specify the desired behavior of the code.  
Verification at the next level, ``platinum level,'' consists verifying a complete set of specifications.
At the moment, the SPARK code generated by Salty proves at gold level.
In particular, SPARK is able to prove the system initial and system transition portions the original GR(1) specification. 
Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
Platinum level would be reached if the liveness properties were expressed and proved, 
which we leave for future work.

@@

In SPARK, subunits consist of a specification and a body.
% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
% (possibly violating the controller's GR(1) specification). 
In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
current input value(s), and current output value(s). 
The internal state number is always initialized to the last possible state number; 
this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition out of this state are those allowed by $\varphi_i^e$.
In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since \lstinline{Controller} is a Moore machine, for each internal state, 
there is exactly one set of output values produced in that state. 
Furthermore, recall that controllers synthesized from GR(1) specifications have the additional property that 
there is exactly one set of input values that brings the controller into each state. 
\lstinline{Controller} therefore includes a type invariant (lines 36-37) that captures this property, 
where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} evaluate 
whether the controller's state corresponds to the expected input and output value(s), respectively. 
% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
unless it is specified during compilation that they should be.
Executing ghost code is often used for debugging, e.g. to check an unproved property through testing. 
In this case, removing the ghost code can save significant memory, 
since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
internally encode large lookup tables that have to store information on every state of the controller, 
which is often thousands of states or even millions of states for the largest example in our database.
The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False}, and 
\lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
since \lstinline{tick = False} brings the system into state 1, and $red$ is true and $yellow$ and $green$ are false in state 1.

@@

We have chosen to describe this particular example in detail because during the process of extending Salty to produce SPARK translations, 
analysis with SPARK revealed a previously undetected problem with this example's specification. 
As written, the specification is realizable and produces what appears at a glance to be a reasonable controller with 97 states. 
In fact, we had previously run this example with OpenAMASE, OpenUxAS, and a Python controller synthesized by Salty. 
However, we did not originally notice in the Python implementation that 34 of the controller's 97 states do not have successors. 
Since no special logic was added to Salty to handle this situation, the generated SPARK code included empty case statement alternatives in the \lstinline{Move} procedure. 
For example, in state 2 the case statement alternative is simply \lstinline{when 2 =>}, with no statements in the body. 
This code failed to compile, since there is no fall-through behavior in case statements in SPARK; 
explicit statements are expected for each case statement alternative. 
In Python, this error went undetected; 
the transition relation for the controller is encoded as a map, and entries corresponding to states without successors 
simply had an empty list of ``next state'' values. 
If the system had attempted to transition states based on an empty list, the result would have been an unhandled runtime error.

@@

Analysis with SPARK did reveal some unexpected behaviors. 
As discussed in \sectref{sec:vipExample}, there were two example controllers\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} 
with specifications that resulted in states with no successors. 
As a result, these controllers contained empty case statements in the \lstinline{Move} procedure. 
We had previously run the Salty-synthesized controller in Python for the example in \sectref{sec:vipExample} and had not noticed the error, 
though it would have resulted in an unhandled runtime exception if one of the states without successors had been reached.

A meta-analysis of SPARK timing results also revealed that other examples in our database did not have any inputs, 
i.e. they amounted to synthesizing a system independent of an environment. 
In those cases, we had specifications for a non-existent environment that were vacuous, 
and this was causing SPARK to take an abnormally long amount of time to verify these controllers, given their relatively small size. 
These controllers did not have errors per se, 
but they were inefficiently encoded. 
We plan to modify Salty to handle such cases by removing the environment, functions over the environment, and
all references to the environment in all pre- and postconditions.
This greatly decreases verification time and also reduces the size and increases the efficiency of the code.

>>>>

SPARK, which is based on the Ada programming language, is both a programming language 
with a specification language and an associated verification toolset \cite{hoang2015spark}.
% SPARK translates code to the Why language for processing by the Why3 theorem prover, 
% and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
For example, if one were to translate source code originally written in Ada to SPARK, 
one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
Formal verification of user-specified properties also requires manually writing specifications at the level of the source code in the 
form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by individual subprograms.
There are different levels to which one can use SPARK to verify a program \cite{moy2019climbing}. 
Generally, the level of verification performed during development is incremental, going from lowest to highest. 
Verification at the lowest level, colloquially referred to as ``stone level,'' is achieved when code is accepted by SPARK, 
since the SPARK language has stricter legality rules than Ada. 
Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
flow analysis checks for errors such as reads of uninitialized data or violations of user-specified data flow contracts. 
Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
e.g. division by zero or numeric overflow/underflow. 
Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
type invariants and subprogram preconditions and postconditions; 
however, at this level, the specifications only partially specify the desired behavior of the code.  
Verification at the next level, ``platinum level,'' consists of verifying a complete set of specifications.
At the moment, the SPARK code generated by Salty proves at gold level.
In particular, SPARK is able to prove the system initial and system transition portions the original GR(1) specification. 
Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
Platinum level would be reached if the liveness properties were expressed and proved, 
which we leave for future work.

@@

In SPARK, subunits consist of a specification and a body.
% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
% (possibly violating the controller's GR(1) specification). 
In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
current input value(s), and current output value(s). 
The internal state number is always initialized to the last possible state number; 
this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition(s) out of this state are those allowed by $\varphi_i^e$.
In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since \lstinline{Controller} is a Moore machine, for each internal state, 
there is exactly one set of output values produced in that state. 
Furthermore, recall that controllers synthesized from GR(1) specifications have the additional property that 
there is exactly one set of input values that brings the controller into each state. 
\lstinline{Controller} therefore includes a type invariant (lines 36-37) that captures this property, 
where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} evaluate 
whether the controller's state corresponds to the expected input and output value(s), respectively. 
% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
unless it is specified during compilation that they should be.
Executing ghost code is often used for debugging, e.g. to check an unproved property through testing. 
In this case, removing the ghost code can save significant memory, 
since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
internally encode large lookup tables that have to store information on every state of the controller, 
which is often thousands of states or even millions of states for the largest example in our database.
The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False}, and 
\lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
since \lstinline{tick = False} brings the system into state 1, and $red$ is true and $yellow$ and $green$ are false in state 1.

@@

We have chosen to describe this particular example in detail because during the process of extending Salty to produce SPARK translations, 
we discovered a previously undetected problem with this example's specification. 
As written, the specification is realizable and produces what appears at a glance to be a reasonable controller with 97 states. 
In fact, we had previously run this example with OpenAMASE, OpenUxAS, and a Python controller synthesized by Salty. 
However, we did not originally notice in the Python implementation that 34 of the controller's 97 states do not have successors. 
Since no special logic was added to Salty to handle this situation, the generated SPARK code included empty case statement alternatives in the \lstinline{Move} procedure. 
For example, in state 2 the case statement alternative is simply \lstinline{when 2 =>}, with no statements in the body. 
This code failed to compile, since SPARK does not allow for fall-through behavior in case statements (nor does Ada); 
explicit statements are expected for each case statement alternative. 
In Python, this error went undetected; 
the transition relation for the controller is encoded as a map, and entries corresponding to states without successors 
simply had an empty list of ``next state'' values. 
We briefly note that if we had encoded the controller using a map in SPARK, 
the error would still have been detected through SPARK analysis rather than a syntactic check of the code; 
the case statement encoding was chosen for efficiency reasons to be discussed in \sectref{sec:lessonsLearned}.

@@

The process of encoding and analyzing controllers in SPARK did reveal some unexpected behaviors. 
First, as discussed in \sectref{sec:vipExample}, there were two example controllers\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} 
with specifications that resulted in states with no successors. 
As a result, these controllers contained empty case statements in the \lstinline{Move} procedure. 
We had previously run the Salty-synthesized controller in Python for the example in \sectref{sec:vipExample} and had not noticed the error, 
though it would have resulted in an unhandled runtime exception if one of the states without successors had been reached.  
Second, a meta-analysis of SPARK timing results also revealed that other examples in our database did not have any inputs, 
i.e. they amounted to synthesizing a system independent of an environment. 
In those cases, we had specifications for a non-existent environment that were vacuous, 
and this was causing SPARK to take an abnormally long amount of time to verify these controllers, given their relatively small size. 
These controllers did not have errors per se, 
but they were inefficiently encoded. 
We plan to modify Salty to handle such cases by removing the environment, functions over the environment, and
all references to the environment in all pre- and postconditions.
This greatly decreases verification time and also reduces the size and increases the efficiency of the code.
